---
title: "Case Study 2: Chulwalar Exports Forecasting Model"
author: "Oscar Padilla"
date: "July 27, 2016"
output:
  html_document:
    keep_md: true
---

# Set Up
```{r}
library(fpp) # for time series forecasting and analysis
library(forecast) # for some other forecasting models
library(Metrics)
library(ggplot2)
setwd("/Users/GS7271/Documents/MSDS6306/Unit10/ChulwalarCase")
```

# Import the data
```{r}
# Import raw .csv files
ImportedAsIsData <- read.csv(file="ImportedAsIsDataChulwalar.csv", header = F, sep=";", fill = T) 
ImportedPlanData <- read.csv(file="ImportedPlanDataChulwalar.csv", header = F, sep=";", fill = T) 
ImportedIndicators <- read.csv(file="ImportedIndicatorsChulwalar.csv", header = F, sep=";", fill = T)
```

# Transform data into time series
In order to be able to work with the partial data sets later, these need to be split into individual vectors and converted into times series.
```{r}
## I would not have this code in the final draft of the case study. These code need to be in an appendix, even if they are put into a makefile.
TotalAsIsVector <- c(ImportedAsIsData [2:13,2],ImportedAsIsData [2:13,3],ImportedAsIsData [2:13,4],ImportedAsIsData [2:13,5],ImportedAsIsData [2:13,6],ImportedAsIsData [2:13,7])
YearAsIsVector <- c(ImportedAsIsData [86,2],ImportedAsIsData [86,3],ImportedAsIsData [86,4],ImportedAsIsData [86,5],ImportedAsIsData [86,6],ImportedAsIsData [86,7])
TotalAsIsVector_2014 <- c(ImportedAsIsData[2:13,8])

PlanVector <- c(ImportedPlanData[2:13,2],ImportedPlanData[2:13,3],ImportedPlanData[2:13,4],ImportedPlanData[2:13,5],ImportedPlanData[2:13,6],ImportedPlanData[2:13,7])
YearPlanVector <- c(ImportedPlanData[86,2],ImportedPlanData[86,3],ImportedPlanData[86,4],ImportedPlanData[86,5],ImportedPlanData[86,6],ImportedPlanData[86,7])
PlanVector_2014 <- c(ImportedPlanData[2:13,8])

# The data is saved as a vector and needs to be converted into a time series
TotalAsIs<- ts(TotalAsIsVector , start=c(2008,1), end=c(2013,12), frequency=12)
YearAsIs <- ts(YearAsIsVector, start=c(2008,1), end=c(2013,12), frequency=12)
TotalAsIs_2014 <- ts(TotalAsIsVector_2014, start=c(2014,1), end=c(2014,12), frequency=12)

TotalPlan <- ts(PlanVector , start=c(2008,1), end=c(2013,12), frequency=12)
YearPlan <- ts(YearPlanVector, start=c(2008,1), end=c(2013,12), frequency=12)
TotalPlan_2014 <- ts(PlanVector_2014, start=c(2014,1), end=c(2014,12), frequency=12)
```

# Basic data analysis

## Use STL function for decomposition
The time series can be analysed using the stl function in order to seperate the trend, seasonality and remainder (remaining coincidential) components from one another.
```{r}
TotalAsIs_stl <- stl(TotalAsIs, s.window=5)
plot(TotalAsIs_stl, col="black", main="TotalAsIs_stl")
plot(TotalAsIs)
lines(TotalAsIs_stl$time.series[,"trend"], col = "red")
```

## Modify seasonal component to a monthly base
The modification of the seasonlity component can also be changed into a monthly view. It only makes sense to do this if the seasonality component as the trend looks almost identical and the remainder is then randomly spread.

```{r}
monthplot(TotalAsIs_stl$time.series[,"seasonal"], main="", ylab="Seasonal")
```

# Correlation with external indicators

## Monthly Change in Export Price Index (CEPI)
```{r}
CEPIVector <- c(ImportedIndicators[2:13,2],ImportedIndicators[2:13,3],ImportedIndicators[2:13,4],ImportedIndicators[2:13,5],ImportedIndicators[2:13,6],ImportedIndicators[2:13,7])
CEPI <- ts(CEPIVector , start=c(2008,1), end=c(2013,12), frequency=12)
plot(CEPI, main="CEPI")
plot(TotalAsIs_stl$time.series[,"trend"]/20000, type = "o", col = "red", ylab = "Trend")
cor(TotalAsIs, CEPI)
cor(TotalAsIs_stl$time.series[,"trend"], CEPI)
```

## Average monthly temperatures in Chulwalar
```{r}
TemperatureVector <- c(ImportedIndicators[30:41,2],ImportedIndicators[30:41,3],ImportedIndicators[30:41,4],ImportedIndicators[30:41,5],ImportedIndicators[30:41,6],ImportedIndicators[30:41,7])
Temperature <- ts(TemperatureVector, start=c(2008,1), end=c(2013,12), frequency=12)
plot(Temperature, main="Temperature")
lines(5 + TotalAsIs_stl$time.series[,"seasonal"]/100000, type ="o", col = "red", ylab = "Seasonality")
cor(TotalAsIs, Temperature)
cor(TotalAsIs_stl$time.series[,"seasonal"], 1/Temperature)
```

## Monthly Average Export Price Index for Chulwalar
```{r}
AEPIVector <- c(ImportedIndicators[100:111,2],ImportedIndicators[100:111,3],ImportedIndicators[100:111,4],ImportedIndicators[100:111,5],ImportedIndicators[100:111,6],ImportedIndicators[100:111,7])
AEPI <- ts(AEPIVector, start=c(2008,1), end=c(2013,12), frequency=12)
plot(AEPI, main="AEPI")

cor(TotalAsIs, AEPI)
```

## National Holidays
```{r}
NationalHolidaysVector <- c(ImportedIndicators[170:181,2],ImportedIndicators[170:181,3],ImportedIndicators[170:181,4],ImportedIndicators[170:181,5],ImportedIndicators[170:181,6],ImportedIndicators[170:181,7])
NationalHolidays <- ts(NationalHolidaysVector, start=c(2008,1), end=c(2013,12), frequency=12)
plot(NationalHolidays, main="NationalHolidays")

cor(TotalAsIs, NationalHolidays)
```

# Monthly Inflation rate in Chulwalar 
```{r}
InflationVector <- c(ImportedIndicators[142:153,2],ImportedIndicators[142:153,3],ImportedIndicators[142:153,4],ImportedIndicators[142:153,5],ImportedIndicators[142:153,6],ImportedIndicators[142:153,7])
Inflation <- ts(InflationVector, start=c(2008,1), end=c(2013,12), frequency=12)
plot(Inflation, main="Inflation")

cor(TotalAsIs, Inflation)
```

# Proposed spending for Independence day presents
```{r}
IndependenceDayPresentsVector <- c(ImportedIndicators[156:167,2],ImportedIndicators[156:167,3],ImportedIndicators[156:167,4],ImportedIndicators[156:167,5],ImportedIndicators[156:167,6],ImportedIndicators[156:167,7])
IndependenceDayPresents <- ts(IndependenceDayPresentsVector, start=c(2008,1), end=c(2013,12), frequency=12)
plot(IndependenceDayPresents, main="IndependenceDayPresents")

cor(TotalAsIs, IndependenceDayPresents)
```


# Forecasting models with smoothing and related approaches
Exponential Smoothing uses past values to calculate a forecast. The strength with which each value influences the forecast is weakened with help of a smoothing parameter. Thus we are dealing with a weighted average, whose values fade out the longer ago they were in the past.

The Akaike's Information Criterion(AIC/AICc) or the Bayesian Information Criterion (BIC) should be at minimum.

## Simple expontential smoothing    
```{r}
Model_ses <- ses(TotalAsIs, h=12)
summary(Model_ses)
plot(Model_ses, plot.conf=FALSE, ylab="Exports Chulwalar", xlab="Year", main="Simple Exponential Smoothing", fcol="white")
lines(fitted(Model_ses), col="green", type="o")
lines(Model_ses$mean, col="blue", type="o")
lines(TotalAsIs_2014, col="red")
legend("topleft", lty = 1, col = c(1,"green"), c("data", expression(alpha == 0.671)), pch=1)

# Now we will compare our model against actual 2014 figures
Model_ses_df <- as.data.frame(Model_ses) 
SES <- ts(Model_ses_df$"Point Forecast", start=c(2014,1), end=c(2014,12), frequency=12)
compareModels <- ts.union(TotalAsIs_2014, SES)
RMSE_SES <- rmse(TotalAsIs_2014, SES)
RMSE_SES
```

## Holt's linear trend method   
Holt added to the model in order to forecast using trends as well. For this it is necessary to add a beta, which determines the trend. If neither alpha nor beta is stated, both parameters will be optimized using ets(). The trend is exponential if the intercepts(level) and the gradient (slope) are multiplied with each other. The values are worse. As the Beta was very low in the optimization, the forecast is very similar to the ses() model. 
```{r}
Model_holt_1 <- holt(TotalAsIs,h=12)
summary(Model_holt_1)
plot(Model_holt_1)
lines(TotalAsIs_2014, col="red")

# Now we will compare our model against actual 2014 figures
Model_holt1_df <- as.data.frame(Model_holt_1) 
Holt_Linear <- ts(Model_holt1_df$"Point Forecast", start=c(2014,1), end=c(2014,12), frequency=12)
compareModels <- ts.union(compareModels, Holt_Linear)
RMSE_LINEAR <- rmse(TotalAsIs_2014, Holt_Linear)
RMSE_LINEAR

# expoential trend
Model_holt_2 <- holt(TotalAsIs, exponential=TRUE,h=12)
summary(Model_holt_2)
plot(Model_holt_2)
lines(TotalAsIs_2014, col="red")

# Now we will compare our model against actual 2014 figures
Model_holt2_df <- as.data.frame(Model_holt_2) 
Holt_Exponential <- ts(Model_holt2_df$"Point Forecast", start=c(2014,1), end=c(2014,12), frequency=12)
compareModels <- ts.union(compareModels, Holt_Exponential)
RMSE_EXPO <- rmse(TotalAsIs_2014, Holt_Exponential)
RMSE_EXPO
```

## Dampened trends
As such simple trends tend to forecast the future to positively, we have added a dampener. This also works for exponential trends. We also plot the level and slope individually for each model.
```{r}
Model_holt_3 <- holt(TotalAsIs, damped=TRUE,h=12)
summary(Model_holt_3)
plot(Model_holt_3)
lines(TotalAsIs_2014, col="red")

# Now we will compare our model against actual 2014 figures
Model_holt3_df <- as.data.frame(Model_holt_3) 
Damped <- ts(Model_holt3_df$"Point Forecast", start=c(2014,1), end=c(2014,12), frequency=12)
compareModels <- ts.union(compareModels, Damped)
RMSE_DAMPED <- rmse(TotalAsIs_2014, Damped)
RMSE_DAMPED

Model_holt_4 <- holt(TotalAsIs, exponential=TRUE, damped=TRUE,h=12)
summary(Model_holt_4)
plot(Model_holt_4)
lines(TotalAsIs_2014, col="red")

# Now we will compare our model against actual 2014 figures
Model_holt4_df <- as.data.frame(Model_holt_4) 
DampedExpo <- ts(Model_holt4_df$"Point Forecast", start=c(2014,1), end=c(2014,12), frequency=12)
compareModels <- ts.union(compareModels, DampedExpo)
RMSE_DAMPEXPO <- rmse(TotalAsIs_2014, DampedExpo)
RMSE_DAMPEXPO

# level and slope can be plotted individually for each model. 
plot(Model_holt_1$model$state)
plot(Model_holt_2$model$state)
plot(Model_holt_3$model$state)
plot(Model_holt_4$model$state)

plot(Model_holt_1, plot.conf=FALSE, ylab="Exports Chulwalar", xlab="Year", main="", fcol="white")
lines(fitted(Model_ses), col="purple", type="o")
lines(fitted(Model_holt_1), col="blue", type="o")
lines(fitted(Model_holt_2), col="red", type="o")
lines(fitted(Model_holt_3), col="green", type="o")
lines(fitted(Model_holt_4), col="orange", type="o")
lines(Model_ses$mean, col="purple", type="o")
lines(Model_holt_1$mean, col="blue", type="o")
lines(Model_holt_2$mean, col="red", type="o")
lines(Model_holt_3$mean, col="green", type="o")
lines(Model_holt_4$mean, col="orange", type="o")
legend("topleft",lty=1, col=c(1,"purple","blue","red","green","orange"), c("data", "SES","Holts auto", "Exponential", "Additive Damped", "Multiplicative Damped"),pch=1)
```

## Holt-Winter's seasonal method   
Holt and Winters have expanded Holt's model further to include the seasonality aspect. The parameter gamma, which is for smoothing the seasonality, was added to achieve this. The values are better than the models without seasonality. This is logical, since the data is strongly influenced by seasonality.  In the following model, none of the parameters are given so that they will be optimised automatically. There are two models: one using an additive error model method and one using a multiplicative error model. The additive model gives slightly better results than the multiplicative model.

```{r}
Model_hw_1 <- hw(TotalAsIs ,seasonal="additive",h=12)
summary(Model_hw_1)
plot(Model_hw_1)
lines(TotalAsIs_2014, col="red")

# Now we will compare our model against actual 2014 figures
Model_hw_1_df <- as.data.frame(Model_hw_1) 
HW_Add <- ts(Model_hw_1_df$"Point Forecast", start=c(2014,1), end=c(2014,12), frequency=12)
compareModels <- ts.union(compareModels, HW_Add)
RMSE_HW_ADD <- rmse(TotalAsIs_2014, HW_Add)
RMSE_HW_ADD

Model_hw_2 <- hw(TotalAsIs ,seasonal="multiplicative",h=12)
summary(Model_hw_2)
plot(Model_hw_2)
lines(TotalAsIs_2014, col="red")

# Now we will compare our model against actual 2014 figures
Model_hw_2_df <- as.data.frame(Model_hw_2) 
HW_Mul <- ts(Model_hw_2_df$"Point Forecast", start=c(2014,1), end=c(2014,12), frequency=12)
compareModels <- ts.union(compareModels, HW_Mul)
RMSE_HW_MUL <- rmse(TotalAsIs_2014, HW_Mul)
RMSE_HW_MUL
```

# Use accuracy measurements to determine which method is best
##Performance comparison of the forecasting models
```{r}
colnames(compareModels) <- c("TotalAsIs2014", "SES", "Holt_Linear", "Holt_Exponential", "Damped", "DampedExpo", "HW_Add", "HW_Mul")
compareModels

plot(Model_hw_1, ylab = "Exports Chulwalar", plot.conf=FALSE, fcol="white", xlab="Year")
lines(fitted(Model_hw_1), col="blue", lty=2)
lines(fitted(Model_hw_2), col="green", lty=2)
lines(Model_hw_1$mean, type="o", col="blue")
lines(Model_hw_2$mean, type="o", col="green")
lines(TotalAsIs_2014, col = "red")
legend("topleft",lty=1, pch=1, col=1:3, c("data","Holt Winters' Additive","Holt Winters' Multiplicative"))

# Output instruction for the data export of the results for further use in Excel.
write.csv(compareModels, file='ChulwalarForecastModelsComparison.csv')
```

# Summarize results
```{r}
rmse_labels <- c("SES", "Holt_Linear", "Holt_Exponential", "Damped", "DampedExpo", "HW_Add", "HW_Mul")
rmse_values <- c(RMSE_SES, RMSE_LINEAR, RMSE_EXPO, RMSE_DAMPED, RMSE_DAMPEXPO, RMSE_HW_ADD, RMSE_HW_MUL)
RMSE_table <- data.frame(rmse_labels, rmse_values)
RMSE_table
g <- ggplot(RMSE_table, aes(rmse_labels, rmse_values))
g + geom_bar(stat = "identity") + scale_fill_grey()
```

Holt-Winter's Multiplicative model has the lowest RMSE (235296.6) when measured against historical (test) data.
However it's only the second best peformance against 2014 actual data 311363.4
Holt-Winter's Additive model is the best performer predicting 2014 data (RMSE = 271816.7)